---
title: 'Google DA Capstone: Cyclistic'
author: "Santiago Rodarte"
date: "2024-04-19"
output: html_document
---

# Google Data Analytics Capstone Project: Cyclistic

For this project, I analyzed the fictional company Cyclistic, where I was tasked 
with completing an analysis of their trip data. The ultimate goal of Cyclistic 
is to convert casual users to members, and this analysis is the first step towards reaching that goal.

I was tasked with the following:

**How do annual members and casual riders use Cyclistic bikes differently?**

This report includes: 

* A brief description of the data sources used
* Documentation on the cleaning process
* Documentation of the analysis

*Key findings, conclusions and recommendations are not included in this report, as they will be included in the presentation*

# Data Used

The data that was used was retrieved from <https://divvybikes.com/system-data>. These datasets were produced by Divvy, a Lyft bike company. It is a repository with historical trip data open to the public. 

After reviewing the data I believe that it is reliable, original, comprehensive, current and cited. 

The datasets that I downloaded form this repository are:

* `Divvy_Trips_2019_Q2.csv`
* `Divvy_Trips_2019_Q3.csv`
* `Divvy_Trips_2019_Q4.csv`
* `Divvy_Trips_2020_Q1.csv`

# Cleaning Process

## Loading the data
Once I had downloaded the data, it was time to load it onto R and begin the exploratory analysis.

```{r loading data}
# Define the filepath tha I will use for all of my datasets
filepath <- ('C:/Users/santi/Desktop/Personal/Cursos/Google Data Analytics/Capstone/Trips Data/')

# Loading the datasets
q2_2019 <- read.csv(paste(filepath,'Divvy_Trips_2019_Q2.csv',sep=''))
q3_2019 <- read.csv(paste(filepath,'Divvy_Trips_2019_Q3.csv',sep=''))
q4_2019 <- read.csv(paste(filepath,'Divvy_Trips_2019_Q4.csv',sep=''))
q1_2020 <- read.csv(paste(filepath,'Divvy_Trips_2020_Q1.csv',sep=''))
```

## Exploring the data
For each one of the datasets, I checked the columns, the types of values, and the dimension, before deciding how I was going to approach the data cleaning process.

```{r exploring datasets}
# Exploring the 2019 Q2 trips dataset
View(q2_2019)
colnames(q2_2019)
str(q2_2019)

# Exploring the 2019 Q3 trips dataset
View(q3_2019)
colnames(q3_2019)
str(q3_2019)

# Exploring the 2019 Q4 trisp dataset
View(q4_2019)
colnames(q4_2019)
str(q3_2019)

#Exploring the 2020 Q1 trips dataset
View(q1_2020)
colnames(q1_2020)
str(q1_2020)
```

## Cleaning the data
After taking a look at the datasets: the 2019 datasets have the same columns, Q1 has different names,so names will have to be changed for all 4 datasets to have the same names. 

1. Names will have to be changed in order for all 4 tables to have the same column names.

2. Characters will have to be changed: id to `chr` for 2019 datasets (2020 ds id columns cannot be changed to int), dates in every tables will be changed to `datetime`, `tripduration` will be deleted and later created into the merged dataset.

3. New columns: 2020 tables will have new column for trip duration, columns will be reordered.

4. Delete columns - in 2019 tables: `bike_id` (there is no `bike_id` in 2020), `gender`, `birthyear` - in 2020 table: `rideable_type`, `start_lat`, `start_lng`, `end_lat`, `end_lng`.

5. 2019 datasets have 'Customer' or 'Subscriber', we will change it to 2020 values 'Casual' or 'Member'

```{r}
# Changing the names for q2_2019
colnames(q2_2019) <- c('trip_id','start_time','end_time', 'bikeid','tripduration',
                       'from_station_id','from_station_name','to_station_id',
                       'to_station_name','usertype','gender','birthyear')
View(q2_2019)

# Now that the 2019 tables have the same names, I will merge them
trips_2019 <- bind_rows(q2_2019,q3_2019,q4_2019)
View(trips_2019)

# I will now remove the columns from the 2019 dataset
trips_2019_v2 <- subset(trips_2019, select = -c(bikeid,tripduration,gender,birthyear))
View(trips_2019_v2)
colnames(trips_2019_v2) <- c('trip_id','start_time','end_time','start_station_id',
                             'start_station_name','end_station_id','end_station_name',
                             'usertype')


# I will now delete the columns from the 2020 table and change the names of the
# remaining columns
q1_2020_v2 <- subset(q1_2020, select = -c(rideable_type, start_lat, start_lng,
                                          end_lat, end_lng))
View(q1_2020_v2)

# Rename columns in 2020 dataset and reorder for the to make sense with 2019 trips
colnames(q1_2020_v2) <- c('trip_id','start_time','end_time','start_station_name','start_station_id',
                          'end_station_name','end_station_id','usertype')
col_order <- c('trip_id','start_time','end_time','start_station_id','start_station_name',
               'end_station_id','end_station_name','usertype')
q1_2020_v3 <- q1_2020_v2[, col_order]

View(q1_2020_v3)


# Prepare to merge the datasets
str(trips_2019_v2)
str(q1_2020_v3)
trips_2019_v2$trip_id <- as.character(trips_2019_v2$trip_id)
str(trips_2019_v2)
```

## Merging the datasets and cleaning final dataset
```{r}
trips_df <- bind_rows(trips_2019_v2,q1_2020_v3)
View(trips_df)

dim(trips_df)
str(trips_df)
summary(trips_df)
```

Steps to take in this part
(1) Change data type of stat time and end time to datetime
(2) Create the trip_duration column using the end_time - start_time
(3) Change 'Customer' to 'casual' and 'Subscriber' to 'member'
(4) Add Day, Month, Year columns
(5) Check for Null values, numerical incoherences, etc.

Convert start time and end time to datetime:
```{r}
trips_df$start_time <- ymd_hms(trips_df$start_time)
trips_df$end_time <- ymd_hms(trips_df$end_time)
str(trips_df)
dim(trips_df)
```

Add new `tripduration` column
```{r}
trips_df$tripduration <- trips_df$end_time - trips_df$start_time
View(trips_df)
str(trips_df)
trips_df$tripduration <- as.numeric(trips_df$tripduration)
dim(trips_df)
```

Add Year, Month, Day, and day of week columns using `lubridate`
```{r}
trips_df$year <- year(trips_df$start_time)
trips_df$month <- month(trips_df$start_time)
trips_df$day <- day(trips_df$start_time)
trips_df$day_of_week <- wday(trips_df$start_time, label=TRUE)
```

Change the subscriber/customer to member/casual
```{r}
trips_df <- trips_df %>%
  mutate(usertype = recode(usertype,
                           'Subscriber'='member',
                           'Customer'='casual'))
table(trips_df$usertype)
```

Now I will delete the rows that have a tripduration < 0:
```{r}
temp_df <- trips_df %>% filter(tripduration < 0)
dim(temp_df)
View(temp_df)

rows_to_remove <- as.vector(temp_df$tripduration)

trips_df_v2 <- trips_df[!(trips_df$start_station_name=='HQ QR' | trips_df$tripduration<0),]
```

 Lastly, I will fix the only null value there is. Since it it a row without info, it will be deleted.
```{r}
trips_df_v3 <- trips_df_v2[complete.cases(trips_df_v2),]
summary(trips_df_v3)
```

Change the tripduration to minutes so it is easier to understand.
```{r}
trips_df_v3 <- trips_df_v3 %>%
  mutate(tripduration = tripduration / 60)
```

# Analyzing process

I will start by analyzing the time duration of the different trips, and how they change depending on the type of user.

```{r}
summary(trips_df_v3$tripduration)

aggregate(trips_df_v3$tripduration~trips_df_v3$usertype, FUN=mean)
aggregate(trips_df_v3$tripduration~trips_df_v3$usertype, FUN=median)
aggregate(trips_df_v3$tripduration~trips_df_v3$usertype, FUN=max)
aggregate(trips_df_v3$tripduration~trips_df_v3$usertype, FUN=min)

aggregate(trips_df_v3$tripduration~trips_df_v3$usertype +
            trips_df_v3$day_of_week, FUN = mean)

trips_df_v3 %>% group_by(usertype,day_of_week) %>% 
  summarise(number_of_rides =n(),
            average_duration = mean(tripduration)) %>% 
  arrange(usertype, day_of_week)
```

Next, I will create visualizations that will allow me to see how the behavior changes depending on the different type of user.

### Average trip duration by user
I will see how the duration of the trip changes for each type of user.
```{r}
trips_df_v3 %>% 
  filter(tripduration<100) %>% 
  ggplot(aes(x=tripduration, fill=usertype)) + 
  geom_density(alpha=0.8) +
  scale_fill_manual(values = c("#0ADD08", "#2290AD")) +
  labs(x='Trip Duration (Minutes)',fill='User Type',
       title='Average Trip Duration by User Type') +
  theme_classic() +
  theme(legend.position = 'bottom')
```

### Number of rides per day of the week
I will see how the number of rides changes throughout the week, and how different types of users behave each day.
```{r}
trips_df_v3 %>% group_by(usertype,day_of_week) %>% 
  summarise(number_of_rides =n(),
            average_duration = mean(tripduration)) %>% 
  arrange(usertype, day_of_week) %>% 
  ggplot(aes(x=day_of_week,y=number_of_rides,fill=usertype)) +
  geom_col(position='dodge') + 
  scale_fill_manual(values = c("#0ADD08", "#2290AD")) +
  theme_classic() +
  theme(legend.position='bottom') +
  labs(x='Weekday',y='Number of Rides',fill='User Type',
       title='Number of Rides per Day of the Week by User Type')
```

**For the next plots, I will have to modify how month are presented in the dataset. Tey are represented as numbers, and I want them to be represented with names.**

```{r}
trips_df_v3$month <- month(trips_df_v3$month, label=TRUE)
```

### Distribution of rides through the year
I will see what percentage of rides are taken by each type of user through the year.
```{r}
trips_df_v3 %>% group_by(usertype, month, year) %>% 
  summarise(number_of_rides= n()) %>%
  filter(number_of_rides > 0) %>% 
  ggplot(aes(x=month,y=number_of_rides,fill=usertype)) +
  geom_bar(position='fill',stat='identity', width=0.7) + 
  facet_wrap(~year, nrow=1) +
  labs(x='Month', y='Proportion of Rides',fill='User Type',
       subtitle='Proportion of the rides per type of user') +
  ggtitle('Distribution of Rides per User Type for the Past Year') +
  theme_classic() +
  theme(legend.position='bottom',axis.text.x = element_text(angle=45)) +
  scale_fill_manual(values = c("#0ADD08", "#2290AD"))
```

### Number of rides through the year
I will see how the number of total rides changes throughout the year, and the number of rides that each user takes per month.
```{r}
trips_df_v3 %>% group_by(usertype, month, year) %>% 
  summarise(number_of_rides= n()) %>% 
  ggplot(aes(x=month,y=number_of_rides,fill=usertype)) +
  geom_bar(position='stack',stat='identity') + 
  facet_wrap(~year,nrow=1) +
  labs(x='Month', y='Number of Rides',fill='User Type',
       subtitle='Count of the rides per type of user') +
  ggtitle('Number of Rides per User Type for the Past Year') +
  theme_classic() +
  theme(legend.position='bottom',axis.text.x = element_text(angle=45)) +
  scale_fill_manual(values = c("#0ADD08", "#2290AD"))
```


# Conclusion
This is the process that I followed for my analysis, from the data loading, to the data cleansing process, to the analysis process. 

Conclusions and recommndations are inside the presentation that i have prepared.



